---
title: "Reading TIVITA recordings with tivis.r"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reading TIVITA recordings with tivis.r}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(tivis.r)
```

The Diaspective Vision TIVITA Suite writes each capture as a directory
containing a binary `*_SpecCube.dat` cube, a `*_meta.log` acquisition record,
and a set of PNG parameter maps the Suite computed itself. `tivis.r` reads all
three. It needs no vendor SDK and contains no compiled code, which is the main
practical difference from its Cubert counterpart, `cuvis.r`.

Everything below runs against a small example recording bundled with the
package, so it works without access to clinical data.

```{r}
f <- tivis_example_file()
basename(f)
```

## Inspecting before reading

A full TIVITA cube is roughly 123 MB. When surveying an archive you rarely want
the samples, so read the header alone — it is a 12-byte read regardless of file
size.

```{r}
tivis_read_header(f)
```

The header carries `width`, `height` and `bands`. Note the ordering: the header
is `(width, height, bands)`, but the array you get back is
`(rows, cols, bands)` — height first — to match the convention used by
`hyperspectR::hsi_cube()`.

## Reading a cube

```{r}
cube <- tivis_read_cube(f)
dim(cube)
range(attr(cube, "wavelengths"))
```

The wavelength axis is attached as an attribute. A native TIVITA recording
covers 500–995 nm in 5 nm steps across 100 bands.

```{r}
head(tivis_get_wavelengths(100L))
```

Reading a subset of bands is often enough, for instance when building an RGB
composite:

```{r}
rgb_bands <- tivis_read_cube(f, bands = c(1L, 3L, 5L))
dim(rgb_bands)
attr(rgb_bands, "wavelengths")
```

## Acquisition metadata

The Suite writes an INI-style log with German section names, `latin1`
encoding and comma decimal separators. All three are handled, and values are
coerced to numerics and logicals where that is unambiguous.

```{r}
log_file <- system.file("extdata", "example_meta.log", package = "tivis.r")
meta <- tivis_get_metadata(log_file)
str(meta, max.level = 2)
```

Note that `Intensity Grenzwert`, written as `7,000000`, comes back as the
number `7` rather than a string, while text that merely contains a comma is
left alone.

```{r}
meta$Fremdlichterkennung$`Intensity Grenzwert`
```

## Whole recordings and archives

`tivis_measurement()` bundles the cube with everything stored beside it:

```{r}
m <- tivis_measurement(f)
names(m)
dim(m$cube)
```

For a real archive, `tivis_list_measurements()` walks the tree and returns one
row per recording, reading only headers:

```{r eval = FALSE}
ms <- tivis_list_measurements("/archive/tivita")
nrow(ms)
unique(paste(ms$width, ms$height, ms$bands))
```

## A note on the format

TIVITA does not publish a specification. The layout implemented here was
determined empirically and then validated against the Suite's own
`_RGB-Image.png` exports: correlating each band of a real recording against the
greyscale reference peaks at band 12, i.e. 555 nm, which is where human
photopic luminous efficiency peaks. Competing hypotheses — band-sequential
storage, transposed spatial axes — score near zero.

This matters because the failure mode is silent. `array()` recycles without
complaint, so a reader with the wrong dimensions or sample type still returns a
full-size cube; it is simply noise. `tivis_read_header()` therefore checks the
file size against the header and refuses to guess.

## Where this fits

`tivis.r` is a format reader, deliberately narrow. Cube construction,
preprocessing and tissue indices live in
[`hyperspectR`](https://github.com/CTTIR/hyperspectR); publication-grade
rendering lives in
[`hyperspectaculR`](https://github.com/CTTIR/hyperspectaculR).
